Local Spectral Clustering of Density Upper Level Sets

### 2019-07-03 neurips/local_spectral_density_clustering_arxiv.tex

\begin{theorem}
  \label{thm: consistent_recovery_of_density_clusters}
  Fix $\lambda > 0$, let $\Cset \in \Cbb_f(\lambda)$ be a
  $\kappa$-well-conditioned density cluster (with respect to some $\theta$), and
  additionally assume $f$ satisfies \ref{asmp: C'_bounded_density}. If Algorithm
  \ref{alg: ppr} is well-initialized, there exists a universal constant $c >  0$
  such that if  
  \begin{equation}
    \label{eqn: kappa_ub}
    \kappa(\Cset) \leq c \frac{\lambda_{\sigma}^2r^d
      \nu_d}{\Lambda_{\sigma}\Pbb(\Csig)}, 
  \end{equation}
  then the output set $\Cest \subseteq \Xbf$ is a consistent estimator for
  $\Cset$, in the sense of Definition \ref{def: consistent_density_cluster_estimation}. 
\end{theorem}

\begin{remark}
  We note that the restriction on $\kappa(\Cset)$ imposed by \eqref{eqn:
    kappa_ub} results in a symmetric set difference \smash{$\Delta(\Csig[\Xbf],
    \Cest)$} on the order of $r^d$. In plain terms, we are able to recover a
  density cluster $\Cset$ in the sense of \eqref{eqn:
    consistent_density_cluster_recovery} only when we can guarantee a very small
  fraction of points will be misclassified. This strong condition is the price
  we pay in order to obtain the uniform bound of \eqref{eqn: ppr_gap}. 
\end{remark}

\begin{remark}
  While taking the radius of the neighborhood graph $r \to 0$ as $n \to
  \infty$ (thereby ensuring $G_{n,r}$ is sparse) is computationally
  attractive, the presence of a factor of \smash{$\log^2(1/r)/r$} in
  $\kappa(\Cset)$ unfortunately prevents us from making claims about the
  behavior of PPR in this regime. Although the restriction to a kernel
  function fixed in $n$ is standard for spectral clustering theory
  \citep{schiebinger2015,vonluxburg2008}, it is an interesting question whether
  PPR exhibits some degeneracy over $r$-neighborhood graphs as $r \to 0$, or if
  this is merely looseness in our upper bounds.  
\end{remark}

\sbcomment{You are using $\epsilon$ here and in the future to mean different things. I have
changed this $\epsilon$ to $\varepsilon$.} 
\paragraph{Approximate PPR vector.}  In practice, exactly solving \eqref{eqn:
  ppr_vector} may be too computationally expensive. To address this limitation,
\citet{andersen2006} introduced the \emph{$\varepsilon$-approximate} PPR vector
(aPPR), which we will denote by \smash{$\pbf^{(\varepsilon)}$}. We refer the
curious reader to \citet{andersen2006} for a formal algorithmic definition of
the aPPR vector, and limit ourselves to highlighting a few salient points: the
aPPR vector can be computed in order $\mathcal{O}(1/(\varepsilon\alpha))$ time,
while satisfying the following uniform error bound: 
\begin{equation}
\label{eqn: appr_error}
\textrm{for all $u \in V$}, \quad \pbf(u) - \varepsilon \Dbf_{uu}\leq
\pbf^{(\varepsilon)}(u) \leq \pbf(u).  
\end{equation}
Application of \eqref{eqn: appr_error} within the proofs of Theorems \ref{thm:
  misclassification_rate} and \ref{thm: consistent_recovery_of_density_clusters}
leads to analogous results which hold for \smash{$\pbf^{(\varepsilon)}$}. We
formally state and prove this in Corollary~\ref{cor: appr} in 
the appendix.

\section{Analysis}
\label{sec: analysis}

@@

Before we proceed to the next subsection, we include a formal statement of our claims regarding the a\pprspace vector. 
\begin{corollary}
	\label{cor: appr}
	Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ be a $\kappa$-well-conditioned cluster (with respect to some $\theta$). Choose input parameters $\alpha, r, \vol_0, v$ to be well-initialized in the sense of \eqref{eqn: initialization}, set $\varepsilon = \frac{1}{20 \vol_0}$, and modify Algorithm \ref{alg: ppr} to compute the a\pprspace vector $\pbf^{(\varepsilon)}$ rather than the exact \pprspace vector $\pbf$, with resulting output $\Cest$.
	\begin{enumerate}
		\item If \ref{asmp: valid_region} holds, then \eqref{eqn: misclassification_rate_ub} is still a valid upper bound for the misclassification error of $\Cest$.
		\item If \ref{asmp: C'_bounded_density} and \eqref{eqn: kappa_ub} hold, 
		then $\Cest \subseteq \Xbf$ is a consistent estimator for $\Cset$, in the sense of Definition \ref{def: consistent_density_cluster_estimation}.
	\end{enumerate}
\end{corollary}

\section{Proof of Theorem \ref{thm: misclassification_rate}}
\label{sec: proof_of_misclassification_rate}
In this section, we prove both Theorem \ref{thm: misclassification_rate} and the relevant portions of Corollary \ref{cor: appr}.

Lemma \ref{lem: zhu} is a restatement of Lemma 3.4 of \cite{zhu2013}, translated into notation consistent with the rest of this paper. It is stated with respect to an arbitrary $S \subseteq G = (V,E)$, and a good set $S^g \subseteq S$ with $\vol(S^g; G) \geq \vol(S; G)/2$. \footnote{The sweep cuts of Lemma 3.4 of \cite{zhu2013} are stated directly with respect to $\vol(S;G)$, rather than $\vol_0$; this discrepancy leads to the extra factors of $1 - c$ and $1 + c$ which appear in our bounds as opposed to theirs.}

\begin{lemma}
	\label{lem: zhu}
	Let $\varepsilon \geq 0$, $v \in S^g$, $\alpha \leq \frac{1}{9\tau_{\infty}(G[S])}$, and $\vol_0 \in [1 - c,1 + c] \vol(S;G)$ for $0 < c < 1$. Consider taking sweep cuts of the approximate \pprspace vector $\pbf^{(\varepsilon)} := \pbf^{(\varepsilon)}(v,\alpha; G)$ of the form
	\begin{equation*}
	S_{\beta} := \left\{u \in V: \pbf^{(\varepsilon)} \geq \beta \frac{\deg(u; G)}{\vol_0} \right\}
	\end{equation*}
	For any $\beta < \frac{3}{5(1 + c)}$, and $0 \leq \varepsilon \leq \frac{1}{10} \vol(S;G)$ the following bounds hold:
	\begin{align}
	\vol(S_{\beta} \setminus S; G) & \leq \frac{2\Phi(A;G)}{\alpha \beta(1 - c)} \vol(S; G), \label{eqn: zhu_1}\\
	\vol(S \setminus S_{\beta}; G) & \leq \left(\frac{2\Phi(S;G)}{\alpha (\frac{3}{5} - \beta(1 + c))} + 8\Phi(S;G)\right)\vol(S; G).  \label{eqn: zhu_2}
	\end{align}
\end{lemma}

\paragraph{Proof of Theorem \ref{thm: misclassification_rate}.}

Observe that by Theorem \ref{thm: mixing_time_upper_bound} and the upper bounds $\alpha \leq \frac{1}{9 \tau_u(\theta)}, r < \sigma/2d$ given in \eqref{eqn: initialization}, with probability one as $n \to \infty$, $\alpha \leq \frac{1}{9\tau_{\infty}(G[S])}$. Moreover, 
\begin{enumerate}[\text{(i)}]
	\item we restrict $v \in \Csig[\Xbf]^g$
	\item the output of Algorithm \ref{alg: ppr} $\widehat{C} = S_{\beta}$ for some $1/40 < \beta < 1/11$
	\item by Lemma \ref{lem: ball_bounds_in_probability}, $\vol_0 \in [1/2,3/2]\vol_{n,r}(\Csig[\Xbf])$
	\item the \pprspace vector $\pbf(v,\alpha;G_{n,r})$ is simply an approximate \pprspace vector with $\varepsilon = 0$.
\end{enumerate}

All conditions of Lemma \ref{lem: zhu} are therefore satisfied, and consequently by \eqref{eqn: zhu_1}, \eqref{eqn: zhu_2}, Theorem \ref{thm: conductance_upper_bound}, and the lower bound $\alpha \geq 1/10\tau_u(\theta)$, the following bounds hold with probability tending to 1 as $n \to \infty$:
\begin{equation*}
\vol_{n,r}(\widehat{C} \setminus \Csig[\Xbf]) \leq 1600 \kappa(\Cset) \vol_{n,r}(\Csig[\Xbf]), \quad \textrm{and} \quad  \vol_{n,r}(\Csig[\Xbf] \setminus \widehat{C}) \leq 52 \kappa(\Cset) \vol_{n,r}(\Csig[\Xbf])
\end{equation*}
Noting that
\begin{align*}
\abs{\widehat{C} \setminus \Csig[\Xbf]} & \leq \frac{\vol_{n,r}(\widehat{C} \setminus \Csig[\Xbf])}{\degmin}   \\
\abs{\Csig[\Xbf] \setminus \widehat{C}} & \leq \frac{\vol_{n,r}(\Csig[\Xbf] \setminus \widehat{C})}{\degminwt}, \quad \textrm{and} \\
\abs{\Csig[\Xbf]} & \geq \frac{\vol_{n,r}(\Csig[\Xbf])}{\degmax}
\end{align*}
the claim follows in light of \eqref{eqn: min_degree_bound}, \eqref{eqn: min_cluster_degree_bound} and \eqref{eqn: max_degree_bound} from Lemma \ref{lem: ball_bounds_in_probability}.

\paragraph{Proof of Corollary \ref{cor: appr} misclassification rates.}

The choice $\varepsilon = \frac{1}{20 \vol_0}$, along with Lemma~\ref{lem: ball_bounds_in_probability} and the upper bound on $\vol_0$ given by \eqref{eqn: initialization}, together imply that $\varepsilon < \frac{1}{10}\vol_{n,r}(\Csig[\Xbf])$ with probability tending to one as $n \to \infty$. We may therefore apply Lemma \ref{lem: zhu} with respect to $\pbf^{(\varepsilon)}$. All other aspects of the proof of Theorem \ref{thm: misclassification_rate} may be applied exactly as stated.

>>>>

\begin{theorem}
  \label{thm: consistent_recovery_of_density_clusters}
  Fix $\lambda > 0$, let $\Cset \in \Cbb_f(\lambda)$ be a
  $\kappa$-well-conditioned density cluster (with respect to some $\theta$), and
  additionally assume $f$ satisfies \ref{asmp: C'_bounded_density}. If Algorithm
  \ref{alg: ppr} is well-initialized, there exists a universal constant $c >  0$
  such that if  
  \begin{equation}
    \label{eqn: kappa_ub}
    \kappa(\Cset) \leq c \frac{\lambda_{\sigma}^2r^d
      \nu_d}{\Lambda_{\sigma}\Pbb(\Csig)}, 
  \end{equation}
  then the output set $\Cest \subseteq \Xbf$ is a consistent estimator for
  $\Cset$, in the sense of Definition \ref{def: consistent_density_cluster_estimation}. 
\end{theorem}
\noindent Some remarks are in order: 
 \begin{enumerate}  
 \item We note that the restriction on $\kappa(\Cset)$ imposed by \eqref{eqn:
    kappa_ub} results in a symmetric set difference \smash{$\Delta(\Csig[\Xbf],
    \Cest)$} on the order of $r^d$. In plain terms, we are able to recover a
  density cluster $\Cset$ in the sense of \eqref{eqn:
    consistent_density_cluster_recovery} only when we can guarantee a very small
  fraction of points will be misclassified. This strong condition is the price
  we pay in order to obtain the uniform bound in~\eqref{eqn: ppr_gap}. 
 \item While taking the radius of the neighborhood graph $r \to 0$ as $n \to
  \infty$ (thereby ensuring that the graph $G_{n,r}$ is sparse) is computationally
  attractive, the presence of a factor of \smash{$\log^2(1/r)/r$} in
  $\kappa(\Cset)$ prevents us from making claims about the
  behavior of PPR in this regime. Although the restriction to a kernel
  function fixed in $n$ is standard for spectral clustering theory
  \citep{schiebinger2015,vonluxburg2008}, it is an interesting question whether
  PPR exhibits some degeneracy over $r$-neighborhood graphs as $r \to 0$, or if
  this is merely looseness in our upper bounds.  
\end{enumerate}
\sbcomment{You are using $\epsilon$ here and in the future to mean different things. I have
changed this $\epsilon$ to $\varepsilon$.}

\subsection{Approximate PPR vector} 
In practice, exactly solving the system of equations~\eqref{eqn:
  ppr_vector} to compute the PPR vector 
  may be too computationally expensive. To address this limitation,
\citet{andersen2006} introduced the \emph{$\varepsilon$-approximate} PPR vector
(aPPR), which we will denote by \smash{$\pbf^{(\varepsilon)}$}. We refer the
curious reader to \citet{andersen2006} for a formal algorithmic definition of
the aPPR vector, and limit ourselves to highlighting a few salient points: the
aPPR vector can be computed in order $\mathcal{O}(1/(\varepsilon\alpha))$ time,
while satisfying the following uniform error bound: 
\begin{equation}
\label{eqn: appr_error}
\textrm{for all $u \in V$}, \quad \pbf(u) - \varepsilon \Dbf_{uu}\leq
\pbf^{(\varepsilon)}(u) \leq \pbf(u).  
\end{equation}
For a sufficiently small choice of $\varepsilon$, the 
application of \eqref{eqn: appr_error} within the proofs of Theorems \ref{thm:
  misclassification_rate} and \ref{thm: consistent_recovery_of_density_clusters}
leads to analogous results which hold for \smash{$\pbf^{(\varepsilon)}$}. We
formally state and prove this in Corollary~\ref{cor: appr} in 
Appendix~\ref{sec:appr}.

\section{Analysis Overview}
\label{sec: analysis}

@@

\section{Proof of Theorem \ref{thm: misclassification_rate}}
\label{sec: proof_of_misclassification_rate}
In this section, we prove both Theorem \ref{thm: misclassification_rate}.
The result is in essence a consequence of the bounds in Theorem~\ref{thm: conductance_upper_bound} 
and Theorem~\ref{thm: mixing_time_upper_bound} bounding the normalized cut and mixing time, together with the result of \citet{zhu2013} relating the quality of the PPR output to these quantities, and the result in Lemma~\ref{lem: ball_bounds_in_probability}
that allows us to translate the guarantees on the volume to guarantees on the symmetric set difference.

% and the relevant portions of Corollary \ref{cor: appr}.
We begin with Lemma \ref{lem: zhu}, which is a restatement of Lemma 3.4 of \citet{zhu2013}, translated into notation consistent with the rest of this paper. It is stated with respect to an arbitrary $S \subseteq G = (V,E)$, and a good set $S^g \subseteq S$ with $\vol(S^g; G) \geq \vol(S; G)/2$.\footnote{The sweep cuts of Lemma 3.4 of \cite{zhu2013} are stated directly with respect to $\vol(S;G)$, rather than $\vol_0$; this discrepancy leads to the extra factors of $1 - c$ and $1 + c$ which appear in our bounds as opposed to theirs.}

\begin{lemma}
	\label{lem: zhu}
	Let $\varepsilon \geq 0$, $v \in S^g$, $\alpha \leq \frac{1}{9\tau_{\infty}(G[S])}$, and $\vol_0 \in [1 - c,1 + c] \vol(S;G)$ for $0 < c < 1$. Consider taking sweep cuts of the approximate \pprspace vector $\pbf^{(\varepsilon)} := \pbf^{(\varepsilon)}(v,\alpha; G)$ of the form
	\begin{equation*}
	S_{\beta} := \left\{u \in V: \pbf^{(\varepsilon)} \geq \beta \frac{\deg(u; G)}{\vol_0} \right\}
	\end{equation*}
	For any $\beta < \frac{3}{5(1 + c)}$, and $0 \leq \varepsilon \leq \frac{1}{10} \vol(S;G)$ the following bounds hold:
	\begin{align}
	\vol(S_{\beta} \setminus S; G) & \leq \frac{2\Phi(A;G)}{\alpha \beta(1 - c)} \vol(S; G), \label{eqn: zhu_1}\\
	\vol(S \setminus S_{\beta}; G) & \leq \left(\frac{2\Phi(S;G)}{\alpha (\frac{3}{5} - \beta(1 + c))} + 8\Phi(S;G)\right)\vol(S; G).  \label{eqn: zhu_2}
	\end{align}
\end{lemma}
%\paragraph{Proof of Theorem \ref{thm: misclassification_rate}.}
With this Lemma in place we
observe that by Theorem \ref{thm: mixing_time_upper_bound} and the upper bounds $\alpha \leq \frac{1}{9 \tau_u(\theta)}, r < \sigma/2d$ given in \eqref{eqn: initialization}, with probability one as $n \to \infty$, $\alpha \leq \frac{1}{9\tau_{\infty}(G[S])}$. Moreover, 
\begin{enumerate}
	\item we restrict $v \in \Csig[\Xbf]^g$
	\item the output of Algorithm \ref{alg: ppr} $\widehat{C} = S_{\beta}$, for some $1/40 < \beta < 1/11$
	\item by Lemma \ref{lem: ball_bounds_in_probability}, $\vol_0 \in [1/2,3/2]\vol_{n,r}(\Csig[\Xbf])$
	\item the \pprspace vector $\pbf(v,\alpha;G_{n,r})$ is simply an approximate \pprspace vector with $\varepsilon = 0$.
\end{enumerate}

All conditions of Lemma \ref{lem: zhu} are therefore satisfied, and consequently by \eqref{eqn: zhu_1}, \eqref{eqn: zhu_2}, Theorem \ref{thm: conductance_upper_bound}, and the lower bound $\alpha \geq 1/10\tau_u(\theta)$, the following bounds hold with probability tending to 1 as $n \to \infty$:
\begin{equation*}
\vol_{n,r}(\widehat{C} \setminus \Csig[\Xbf]) \leq 1600 \kappa(\Cset) \vol_{n,r}(\Csig[\Xbf]), \quad \textrm{and} \quad  \vol_{n,r}(\Csig[\Xbf] \setminus \widehat{C}) \leq 52 \kappa(\Cset) \vol_{n,r}(\Csig[\Xbf])
\end{equation*}
Noting that
\begin{align*}
\abs{\widehat{C} \setminus \Csig[\Xbf]} & \leq \frac{\vol_{n,r}(\widehat{C} \setminus \Csig[\Xbf])}{\degmin}   \\
\abs{\Csig[\Xbf] \setminus \widehat{C}} & \leq \frac{\vol_{n,r}(\Csig[\Xbf] \setminus \widehat{C})}{\degminwt}, \quad \textrm{and} \\
\abs{\Csig[\Xbf]} & \geq \frac{\vol_{n,r}(\Csig[\Xbf])}{\degmax}
\end{align*}
the theorem follows in light of \eqref{eqn: min_degree_bound}, \eqref{eqn: min_cluster_degree_bound} and \eqref{eqn: max_degree_bound} from Lemma \ref{lem: ball_bounds_in_probability}.
